---
title: "PM2.5 East and West Europe"
author: "Jusab Aziz (2763065), Angelo Habib (2821154),Abba Ayad (2737337), Chaniro Kocuvi-Tay (2862242), Nasir Ali (2869537), Mikail Cörüz (2863606), Fatima Malik (2788224)"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages, include=FALSE}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(janitor)
require(rnaturalearth)
require(rnaturalearthdata)
require(sf)
```

# Title Page

Jusab Aziz, Angelo Habib, Abba Ayad, Chaniro Kocuvie-Tay, Nasir Ali, Mikail Cörüz, Fatima Malik

Tutorial group 2

Chantal Schouwenaar

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

Air pollution remains one of the greatest environmental risks to human health worldwide, particularly in rapidly growing urban areas across Asia and Europe. Among the various pollutants, fine particulate matter (PM2.5)—particles smaller than 2.5 micrometers—poses the most severe health threat. These microscopic particles can penetrate deep into the lungs and enter the bloodstream, contributing to chronic health conditions such as asthma, cardiovascular disease, lung cancer, and premature death. Neither the World Health Organization (WHO) nor the European Environment Agency (EEA) defines a "safe" threshold for PM2.5 exposure, especially for vulnerable populations such as children, the elderly, and individuals with pre-existing health conditions.

Although progress has been made in EU-wide environmental governance and emissions reduction in Western Europe, regional disparities in air quality persist. While PM2.5 concentrations have declined in much of Western Europe due to improved controls and enforcement, levels remain significantly higher in Eastern Europe. This is largely attributed to continued reliance on coal, high levels of industrial activity, and weaker regulatory enforcement. According to data from the EEA (2024) and Eurostat (2024), urban populations in Eastern Europe are exposed to considerably higher levels of PM2.5 than their counterparts in the West.

These regional imbalances raise critical concerns about environmental inequality and the unequal distribution of health risks. In this paper, we investigate the health impacts of PM2.5 pollution by comparing PM2.5-related mortality rates in Eastern and Western European countries.

Our analysis relies on two main data sources. First, we use average PM2.5 concentration values from urban monitoring stations across European countries, obtained from Eurostat and aggregated at the national level. Second, we use mortality data from the EEA to calculate the number of deaths attributable to PM2.5 exposure per 100,000 people, allowing for comparison across countries of varying population sizes.

By combining these datasets, we aim to provide a clearer understanding of how PM2.5 pollution affects public health across Europe. Highlighting the disparities between East and West underscores the need for targeted and equitable air quality policies aligned with WHO guidelines.

\

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data, include=FALSE}

# Load required packages
library(dplyr)
library(readxl)

east_base_path <- "~/Documents/GitHub/PM2.5/data/east europe data/"

read_tsv_utf16_east <- function(filename, year) {
  path <- paste0(east_base_path, filename)
  df <- read.delim(path, fileEncoding = "UTF-16LE", sep = "\t", check.names = TRUE)
  df$Year <- year
  df <- df[, !duplicated(names(df))]  # remove duplicated columns
  return(df)}

east_2018 <- read_tsv_utf16_east("2018.csv", 2018)
east_2019 <- read_tsv_utf16_east("2019.csv", 2019)
east_2020 <- read_tsv_utf16_east("2020.csv", 2020)
east_2021 <- read_tsv_utf16_east("2021.csv", 2021)
east_2022 <- read_tsv_utf16_east("2022.csv", 2022)
east_2023 <- read_tsv_utf16_east("2023.csv", 2023)

west_base_path <- "~/Documents/GitHub/PM2.5/data/west europe data/"

read_tsv_utf16_west <- function(filename, year) {
  path <- paste0(west_base_path, filename)
  df <- read.delim(path, fileEncoding = "UTF-16LE", sep = "\t", check.names = TRUE)
  df$Year <- year
  df <- df[, !duplicated(names(df))]
  return(df)}

west_2018 <- read_tsv_utf16_west("2018 west.csv", 2018)
west_2019 <- read_tsv_utf16_west("2019 west.csv", 2019)
west_2020 <- read_tsv_utf16_west("2020 west.csv", 2020)
west_2021 <- read_tsv_utf16_west("2021 west.csv", 2021)
west_2022 <- read_tsv_utf16_west("2022 west.csv", 2022)
west_2023 <- read_tsv_utf16_west("2023 west.csv", 2023)

mortality_excel_path <- path.expand("~/Documents/GitHub/PM2.5/data/sdg_11_52_page_spreadsheet-3.xlsx")

df_mortality_raw <- read_excel(mortality_excel_path, skip = 9)
```

**The countries we use for East are:**

Bulgaria, Estonia, Hungary, Latvia, Lithuania, Poland, Romania, Slovakia, Slovenia

**The countries we use for West are:**

Austria, Belgium, France, Germany, Ireland, Luxembourg, Netherlands

### Merging

Here I will merge the two data sets that I have loaded in.

```{r}
all_data_east <- bind_rows(
  east_2018, east_2019, east_2020,
  east_2021, east_2022, east_2023)

all_data_west <- bind_rows(
  west_2018, west_2019, west_2020,
  west_2021, west_2022, west_2023)
```

## 2.2 Provide a short summary of the dataset(s)

```{r include=FALSE}
# Show structure of each raw dataset
head(all_data_east)
head(all_data_west)
```

```{r}
head(df_mortality_raw)
```

The "all_data_east" and "all_data_west" datasets contain yearly air pollution data for each country. Each row represents one country in a specific year (from 2018 to 2023). The variables include PM2.5 concentration, number of monitoring stations, threshold exceedances, and other environmental indicators. A "Year" column was added manually to keep track of the time dimension.

The "df_mortality_raw" dataset shows the estimated number of premature deaths attributed to PM2.5 pollution. It is structured in wide format, with countries in the rows and years (e.g., 2016–2022) as column headers. Each value represents the total number of premature deaths for that country in a given year.

## 2.3 Describe the type of variables included

#### **1. Air Pollution Datasets (all_data_east, all_data_west)**

The [air pollution datasets](https://www.eea.europa.eu/en/analysis/maps-and-charts/air-quality-statistics-dashboards) provide annual, country-level environmental data for the years 2018 to 2023. Most variables are numerical and describe air quality levels, monitoring activity, and regulatory exceedances.

**Key variables include:**

-   **PM2.5 concentration**: The annual average level of fine particulate matter in the air, measured in micrograms per cubic meter (µg/m³).

-   **Number of monitoring stations**: The total number of stations reporting valid data per country and year.

-   **Threshold exceedances**: The number of instances where measured pollution levels exceeded legal EU air quality thresholds.

-   **Population exposure**: In some years, indicators report the proportion of the population exposed to levels above legal limits.

-   **Country**: The name of the reporting country.

-   **Year**: The calendar year of the observation (manually added).

**Data source and collection method:**\
These data are collected through official air quality monitoring stations managed by national environmental authorities. The values are compiled and standardized by the European Environment Agency (EEA). All data are obtained from administrative or sensor-based sources, not from individuals or surveys.

#### **2. Mortality Dataset (df_mortality_raw)**

[The mortality data set contains](https://ec.europa.eu/eurostat/databrowser/view/sdg_11_52/default/table) health outcome estimates related to air pollution exposure. It presents the number of premature deaths attributable to PM2.5 per country and year, based on modelled estimates.

**Key variables include:**

-   **TIME**: Country name.

-   **Year columns**: Each year column represents the total number of estimated premature deaths for that country and year.

**Data source and collection method:**\
This data set originates from the United Nations Sustainable Development Goals (SDG) indicator database, specifically SDG 11.6.2. The figures are not directly observed but are produced using statistical models that combine environmental exposure, population demographics, and epidemiological data. As such, these are administrative and model-based estimates, not derived from individual-level surveys or medical records.

# Part 3 - Quantifying

## 3.1 Data cleaning

```{r data_cleaning, include=FALSE}
# Add Region
all_data_east$Region <- "East"
all_data_west$Region <- "West"

# Ensure both have 'Threshold.1' column
if (!"Threshold.1" %in% colnames(all_data_east)) {
  all_data_east$Threshold.1 <- NA_character_
}
if (!"Threshold.1" %in% colnames(all_data_west)) {
  all_data_west$Threshold.1 <- NA_character_
}
# Force 'Threshold.1' to character in both
all_data_east$Threshold.1 <- as.character(all_data_east$Threshold.1)
all_data_west$Threshold.1 <- as.character(all_data_west$Threshold.1)
# Align both datasets to shared columns
common_cols <- intersect(names(all_data_east), names(all_data_west))
all_data_east <- all_data_east[, common_cols]
all_data_west <- all_data_west[, common_cols]
```

### Merging

```{r}
# Combine
combined_data <- bind_rows(all_data_east, all_data_west)
```

## Data cleaning

Here I will data clean the combined data file because it included a lot of weird numbers and rows.

```{r include=FALSE}
df_combined_clean <- combined_data %>%
  clean_names() %>%
  mutate(across(where(is.character), str_trim)) %>%
  mutate(across(where(is.character), ~na_if(., ""))) %>%
  filter(!if_all(everything(), is.na)) %>%                  
  filter(!is.na(country)) %>%                               
  filter(!grepl("^PM2.5", country))                         
```

## Data cleaning (average of cities PM2.5)

```{r}
# Compute average PM2.5 by country, region, and year
avg_pm25_by_country_year <- df_combined_clean %>%
  group_by(country, region, year_of_statistics) %>%
  summarise(average_pm25 = mean(air_pollution_level, na.rm = TRUE), .groups = "drop") %>%
  arrange(year_of_statistics, region, country)
```

## Data cleaning (mortality)

Here I will data clean the mortality data file because it included a lot of weird numbers.

```{r include=FALSE}
df_mortality_clean <- df_mortality_raw %>%
  rename(country = TIME) %>%
  pivot_longer(
    cols = -country,
    names_to = "year_of_statistics",
    values_to = "premature_deaths",
    values_drop_na = TRUE
  ) %>%
  mutate(
    year_of_statistics = as.integer(year_of_statistics),
    premature_deaths = as.numeric(premature_deaths))
```

## 3.2 Generate necessary variables

### Merging final

```{r}
merged_data <- full_join(df_mortality_clean, avg_pm25_by_country_year, by = c("year_of_statistics", "country"))
```

**Merging to get the filtered data to use further**

```{r}
merged_data$premature_deaths <- as.numeric(gsub(",", "", merged_data$premature_deaths))
filtered_data <- merged_data %>%
  filter(!is.na(premature_deaths), !is.na(average_pm25))
countries_with_data <- unique(filtered_data$country)
```

Variable 1: Population

```{r gen_var1}
population_region_df <- tibble::tibble(
  country = c(
    "Austria", "Belgium", "Bulgaria", "Estonia", "France", "Germany", "Hungary",
    "Ireland", "Latvia", "Lithuania", "Luxembourg", "Netherlands", "Poland",
    "Romania", "Slovakia", "Slovenia"
  ),
  population = c(
    9, 11.5, 6.5, 1.3, 67, 83, 9.6, 5.1, 1.9, 2.8, 0.65, 17, 38, 19, 5.4, 2.1
  ) * 1e6,
  region = c(
    "Western", "Western", "Eastern", "Eastern", "Western", "Western", "Eastern",
    "Western", "Eastern", "Eastern", "Western", "Western", "Eastern", "Eastern",
    "Eastern", "Eastern"))
```

Variable 2: Average deaths

```{r gen_var2}
avg_deaths <- filtered_data %>%
  group_by(country) %>%
  summarise(avg_deaths = mean(premature_deaths, na.rm = TRUE)) %>%
  inner_join(population_region_df, by = "country") %>%
  mutate(
    deaths_per_100k = (avg_deaths / population) * 100000)
```

Variable 3: deaths per 100.000 people

```{r gen_var3}
region_summary <- avg_deaths %>%
  group_by(region) %>%
  summarise(
    avg_deaths_per_100k = mean(deaths_per_100k),
    .groups = "drop"
  ) %>%
  mutate(region = fct_reorder(region, avg_deaths_per_100k))
```

## 3.3 Visualize temporal variation

```{r}
pm25_trend <- merged_data %>%
  filter(!is.na(average_pm25), !is.na(year_of_statistics)) %>%
  group_by(year_of_statistics) %>%
  summarise(avg_pm25 = mean(average_pm25, na.rm = TRUE)) %>%
  ungroup()

# Plot
ggplot(pm25_trend, aes(x = year_of_statistics, y = avg_pm25)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(aes(label = round(avg_pm25, 2)), vjust = -0.8, size = 3.5, fontface = "bold") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Average PM2.5 Over Time in Europe",
    x = "Year",
    y = "Average PM2.5 (µg/m³)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1))
```

The line chart presents average PM2.5 concentrations across Europe from 2018 to 2023. PM2.5, or fine particulate matter, poses serious health risks when levels exceed recommended thresholds. In 2018, Europe recorded the highest concentration in the dataset, at 14.17 µg/m³. This was followed by a gradual decline to 12.1 µg/m³ in 2019 and a more pronounced drop to 10.94 µg/m³ in 2020. The sharp decline in 2020 is likely associated with the COVID-19 pandemic, during which lockdowns significantly reduced industrial activity and transportation.

In 2021, PM2.5 levels rose slightly to 11.56 µg/m³, reflecting the partial return of economic and social activities. However, this increase was not sustained. Concentrations decreased again in 2022 to 11.06 µg/m³ and fell further to 10.13 µg/m³ in 2023—the lowest value in the time series.

Overall, the data indicate a general downward trend in PM2.5 pollution in Europe over the six-year period, with the most significant reduction occurring during the pandemic year.\

## 3.4 Visualize spatial variation

```{r visualise_map, echo=FALSE}
merged_data$premature_deaths <- as.numeric(gsub(",", "", merged_data$premature_deaths))
spatial_data <- merged_data %>%
  filter(!is.na(premature_deaths), !is.na(average_pm25)) %>%
  group_by(country) %>%
  summarise(
    avg_pm25 = mean(average_pm25, na.rm = TRUE),
    avg_deaths = mean(premature_deaths, na.rm = TRUE)
  )
world <- ne_countries(scale = "medium", returnclass = "sf")

# Merge map with your country-level data
map_data <- left_join(world, spatial_data, by = c("name" = "country"))

# Optional: Keep only Europe for clarity
map_data <- map_data %>% filter(region_un == "Europe")
ggplot(map_data) +
  geom_sf(aes(fill = avg_pm25), color = "white", size = 0.2) +
  scale_fill_gradient(
    low = "#c6dbef",
    high = "#08306b",
    na.value = "lightgrey"
  ) +
  labs(
    title = "Spatial Variation of Average PM2.5 in Europe",
    subtitle = "Darker colors indicate higher pollution levels",
    fill = "Avg PM2.5 (µg/m³)",
    caption = "Countries with missing data are shown in grey."
  ) +
  coord_sf(
    xlim = c(-15, 35),
    ylim = c(35, 60),
    expand = FALSE
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10))
```

The choropleth map illustrates the distribution of average PM2.5 concentrations across European countries. Darker shades of blue represent higher concentrations, while lighter shades indicate lower levels. Countries for which data is unavailable are shown in grey.

The map reveals clear regional differences in air pollution levels, with Eastern and Southern European countries generally exhibiting higher PM2.5 concentrations than Western Europe. These geographic disparities highlight the existence of pollution hotspots, supporting the case for region-specific environmental and public health interventions.

A spatial approach such as this is valuable for informing targeted policy responses. It reinforces the need for increased funding and coordinated efforts at the regional level—particularly in areas most affected by elevated air pollution. Tailored strategies addressing the unique industrial, energy, and regulatory challenges in these regions could help mitigate the health risks associated with long-term PM2.5 exposure.3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
ggplot(region_summary, aes(x = region, y = avg_deaths_per_100k, fill = region)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Western" = "steelblue", "Eastern" = "tomato")) +
  labs(
    title = "PM2.5-Related Premature Deaths per 100,000 People",
    subtitle = "Western vs Eastern Europe",
    x = "Region",
    y = "Deaths per 100,000",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 10))
```

This bar chart compares the average PM2.5-related premature death rates per 100,000 people between Eastern and Western Europe. The horizontal layout clearly highlights the regional disparity: Eastern Europe reports an average mortality rate of approximately 80 deaths per 100,000 people or higher, while Western Europe records a significantly lower rate of around 30 deaths per 100,000.

This difference strongly suggests that populations in Eastern Europe are more exposed to the health risks associated with PM2.5 pollution. Several contributing factors may underlie this disparity, including higher levels of pollution from coal-based heating and heavy industry, weaker air quality regulations and enforcement, and less access to high-quality healthcare and pollution mitigation infrastructure.

In summary, the chart underscores a major issue of environmental health inequality within Europe. It reinforces the urgent need for targeted public health and environmental policies in the most affected regions, particularly in Eastern Europe, where the pollution burden and its health consequences are disproportionately high.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis, echo=FALSE}
event_year <- 2020

pm25_event <- merged_data %>%
  group_by(year_of_statistics) %>%
  summarise(avg_pm25 = mean(average_pm25, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(year_of_statistics >= 2018)  # Restrict to 2018 and later

ggplot(pm25_event, aes(x = year_of_statistics, y = avg_pm25)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  

  geom_vline(xintercept = event_year, linetype = "dashed", color = "red", size = 1) +
  
  annotate("text",
           x = event_year,
           y = max(pm25_event$avg_pm25) + 0.5,
           label = "COVID-19 Lockdown in 2020\nLed to a sharp drop in PM2.5 levels\ndue to reduced traffic & industry",
           color = "red",
           fontface = "bold",
           hjust = 0.5,
           size = 4.2) +

  geom_text(aes(label = round(avg_pm25, 2)), vjust = -0.8, size = 3.5, fontface = "bold") +
  
  theme_minimal(base_size = 13) +
  labs(
    title = "Impact of COVID-19 Lockdown on PM2.5 Levels in Europe",
    x = "Year",
    y = "Average PM2.5 (µg/m³)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

This line graph illustrates average PM2.5 levels in Europe between 2018 and 2023, highlighting the effect of the COVID-19 lockdown in 2020. The dashed vertical red line marks the year 2020, when most European countries implemented widespread lockdowns and restrictions.

Prior to 2020, PM2.5 concentrations were already declining—from 14.17 µg/m³ in 2018 to 12.1 µg/m³ in 2019. In 2020, there was a sharp drop to 10.94 µg/m³, the lowest value recorded during the six-year period. This significant decline is closely associated with the lockdown measures, which led to dramatic reductions in transportation, industrial activity, and public movement.

In 2021, PM2.5 levels rose again to 11.56 µg/m³, reflecting the partial resumption of economic and social activities. In 2022, levels slightly declined to 11.06 µg/m³, indicating some continued environmental improvement, though not as pronounced as during the height of the lockdown.

Overall, the graph provides compelling visual evidence of how large-scale shifts in human behavior can influence environmental outcomes. The temporary reduction in air pollution during 2020 demonstrates the potential for policy-driven reductions in emissions through coordinated and sustained action.

## 4.1 Discuss your findings

Our analysis has revealed clear regional disparities in air quality and the associated public health risks across Europe, particularly between Eastern and Western European countries.

**First**, the regional data show that Eastern European countries consistently record higher average PM2.5 concentrations than their Western counterparts. Poland, Romania, Bulgaria, and Hungary had the highest levels of fine particulate matter, while countries like France, Germany, and Ireland showed some of the lowest pollution levels. These findings support the notion that air pollution is geographically uneven across Europe. Contributing factors likely include outdated industrial infrastructure, reliance on coal-based energy, and weaker environmental enforcement in Eastern Europe.

**Second**, the temporal data from 2018 to 2023 indicate a general decline in PM2.5 concentrations, with a particularly sharp drop in 2020—likely due to reduced industrial activity and mobility during the COVID-19 lockdowns. However, the subsequent rebound in 2021 suggests that these improvements were temporary. Without sustained structural changes, reductions in air pollution will be difficult to maintain. This highlights the importance of long-term environmental reforms rather than temporary behavioral shifts prompted by crises.

**Third**, our subgroup analysis underscores significant health disparities. Eastern European countries are not only exposed to higher pollution levels but also suffer from substantially higher rates of premature deaths related to PM2.5 exposure. For instance, Bulgaria, Romania, and Poland report over 150 deaths per 100,000 people, while Western countries such as France, Belgium, and Austria report less than half that figure. This strongly suggests a correlation between pollution exposure and negative health outcomes. It also underscores the need for equitable health and environmental policy responses across Europe.

**In conclusion**, our findings indicate that PM2.5 pollution is both an environmental and social justice issue. The disproportionate burden borne by Eastern European populations calls for urgent and coordinated EU-wide action. This includes investment in clean energy transitions, modernization of industrial infrastructure, and strengthening of public health systems in the most affected regions. Without such measures, pollution-related health disparities will persist, continuing to harm the populations least equipped to manage their effects.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: <https://github.com/angelohbb/PM2.5>

## 5.2 Reference list

European Environment Agency. (2024). *Air quality statistics dashboards*. <https://www.eea.europa.eu/en/analysis/maps-and-charts/air-quality-statistics-dashboards>

Eurostat. (2024). *Urban population exposure to air pollution by particulate matter (PM2.5)*. <https://ec.europa.eu/eurostat/databrowser/view/sdg_11_52/default/table>

World Health Organization. (2024). *Air pollution*. <https://www.who.int/health-topics/air-pollution#tab=tab_1>

\
