---
title: "PM2.5 East and West Europe"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Jusab Aziz, Angelo Habib, Abba Ayad

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

-   Why is this relevant?

-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data, include=FALSE}
# STEP 2.1 — Load in the Starting Data

# Load required packages
library(dplyr)
library(readxl)

# ----------------------------
# Load East Europe Data
# ----------------------------
east_base_path <- "~/Documents/GitHub/PM2.5/data/east europe data/"

read_tsv_utf16_east <- function(filename, year) {
  path <- paste0(east_base_path, filename)
  df <- read.delim(path, fileEncoding = "UTF-16LE", sep = "\t", check.names = TRUE)
  df$Year <- year
  df <- df[, !duplicated(names(df))]  # remove duplicated columns
  return(df)
}

east_2018 <- read_tsv_utf16_east("2018.csv", 2018)
east_2019 <- read_tsv_utf16_east("2019.csv", 2019)
east_2020 <- read_tsv_utf16_east("2020.csv", 2020)
east_2021 <- read_tsv_utf16_east("2021.csv", 2021)
east_2022 <- read_tsv_utf16_east("2022.csv", 2022)
east_2023 <- read_tsv_utf16_east("2023.csv", 2023)

all_data_east <- bind_rows(
  east_2018, east_2019, east_2020,
  east_2021, east_2022, east_2023
)

# ----------------------------
# Load West Europe Data
# ----------------------------
west_base_path <- "~/Documents/GitHub/PM2.5/data/west europe data/"

read_tsv_utf16_west <- function(filename, year) {
  path <- paste0(west_base_path, filename)
  df <- read.delim(path, fileEncoding = "UTF-16LE", sep = "\t", check.names = TRUE)
  df$Year <- year
  df <- df[, !duplicated(names(df))]
  return(df)
}

west_2018 <- read_tsv_utf16_west("2018 west.csv", 2018)
west_2019 <- read_tsv_utf16_west("2019 west.csv", 2019)
west_2020 <- read_tsv_utf16_west("2020 west.csv", 2020)
west_2021 <- read_tsv_utf16_west("2021 west.csv", 2021)
west_2022 <- read_tsv_utf16_west("2022 west.csv", 2022)
west_2023 <- read_tsv_utf16_west("2023 west.csv", 2023)

all_data_west <- bind_rows(
  west_2018, west_2019, west_2020,
  west_2021, west_2022, west_2023
)

# ----------------------------
# Load Mortality (Premature Deaths) Dataset
# ----------------------------
mortality_excel_path <- path.expand("~/Documents/GitHub/PM2.5/sdg_11_52_page_spreadsheet-3.xlsx")

df_mortality_raw <- read_excel(mortality_excel_path, skip = 9)

# Optional: preview all raw datasets
head(all_data_east)
head(all_data_west)
head(df_mortality_raw)

```

## 2.2 Provide a short summary of the dataset(s)

```{r}
# Show structure of each raw dataset
summary(all_data_east)
summary(all_data_west)
head(df_mortality_raw)

```

The air pollution datasets were obtained from the European Environment Agency (EEA) and include data from 2018 to 2023. Each year’s file contains country-level indicators such as:

-   **PM2.5 concentration (µg/m³)**: annual average exposure to fine particulate matter

-   **Number of stations**: count of monitoring stations reporting valid data

-   **Threshold exceedances**: number of days or measurements above EU air quality limits

-   **Urban and rural averages**: in some years, separate data is given for urban/rural zones

The data is measured and reported according to EEA guidelines, but some column names are unclear (`Threshold.1`) and units are missing for certain variables.

The premature mortality dataset is sourced from the UN SDG indicators (Goal 11.6.2). It reports:

-   **Premature deaths**

-   **By country and year**

The definition of premature death reflects deaths attributable to air pollution.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}


```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

### Merging

```{r}
# Combine all datasets
all_data <- bind_rows(
  data_2018, data_2019, data_2020,
  data_2021, data_2022, data_2023
)

# Combine all years
all_data_west <- bind_rows(
  data_2018, data_2019, data_2020,
  data_2021, data_2022, data_2023
)

# Save the combined dataset
write.csv(all_data, "~/Documents/GitHub/PM2.5//All_Years_East_Combined.csv", row.names = FALSE)

# Save the combined West Europe dataset
write.csv(all_data_west, "~/Documents/GitHub/PM2.5/All_Years_Combined_West.csv", row.names = FALSE)

```

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

Our analysis reveals a clear regional divide in air quality and its impact on public health across Europe, specifically when comparing Eastern and Western Europe.

First, the spatial analysis shows that Eastern European countries consistently have higher average PM2.5 levels than Western European countries. Countries like Poland, Romania, Bulgaria, and Hungary showed the most concerning pollution levels. In contrast, France, Germany, and Ireland had much lower PM2.5 concentrations. This supports the hypothesis that air pollution is geographically uneven across Europe, and suggests that factors such as outdated industrial infrastructure, energy production methods, and less stringent environmental policies in the East contribute to this discrepancy.

Second, the temporal analysis from 2018 to 2022 shows a general decline in PM2.5 levels, reaching a low point in 2020. This dip likely correlates with the COVID-19 pandemic, during which lockdowns reduced transportation and industrial emissions. However, the rebound in 2021 indicates that these reductions were temporary, and that without long-term policy changes, improvements in air quality may not be sustained. This emphasizes the need for permanent structural interventions rather than relying on short-term events.

Third, the subgroup analysis highlights the public health consequences of this pollution. Eastern European countries not only have higher PM2.5 levels but also show significantly higher mortality rates related to air pollution. For instance, Bulgaria, Romania, and Poland exhibit more than 150 deaths per 100,000 people, whereas Western European countries like France, Belgium, and Austria report less than half that amount. This strongly suggests that higher pollution directly correlates with higher health risks, and it highlights the urgent need for targeted health and environmental policies in Eastern Europe.

In conclusion, our findings support the idea that PM2.5 pollution is both an environmental and a social issue, with unequal health outcomes across Europe. Tackling this problem requires coordinated policy efforts, especially in Eastern Europe, focusing on clean energy, industrial modernization, and public health investment. Without addressing these disparities, the health impacts of air pollution will continue to disproportionately affect populations in the East.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

World Health Organization. (2021). Air pollution: Health impacts. <https://www.who.int/health-topics/air-pollution#tab=tab_1>

\
